Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2021-05-24 paper_io_sim.tex
Text edits

    The emergence of Big Data in recent years has resulted in a growing
    need for efficient data processing solutions. While infrastructures
    with sufficient compute power are available,
    the I/O bottleneck remains. The Linux page cache is an efficient
    approach to reduce I/O overheads, but few
    experimental studies of its interactions with Big Data applications exist,
    partly due to limitations of
    real-world experiments. Simulation is a popular approach to address
    these issues, however, existing simulation frameworks do not simulate
    page caching fully, or even at all.  As a result, simulation-based
    performance studies of data-intensive applications can lead to misleading
    results.

    In this paper, we propose an I/O simulation model that captures
    the key features of the Linux page cache. We have implemented this model
    as part of the \wrench workflow simulation framework, which itself
    builds on the popular \simgrid distributed systems simulation
    framework. Our model and its implementation enable the simulation
    of both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems. We evaluate the accuracy of our model in different
    conditions, including sequential and concurrent applications, as
    well as local and remote I/Os. We find that our page cache model
    reduces the simulation error by up to an order of magnitude when
    compared to state-of-the-art, cacheless simulations.

@@

        The Linux page cache plays an important role in
        reducing filesystem data transfer times. With the page cache, previously
        read data can be re-read directly from memory, and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance improvements
        depend on many factors including the total amount of memory,
        the amount of data being written (i.e., dirty data), and the amount of memory available for
        written data. All these factors are important when determining the impact of I/O on
        application performance, particularly for data-intensive applications.

        % Swapped the order the two paragraphs and merged them into one
        The number of data-intensive applications has been steadily rising as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large-scale infrastructures
    such as High Performance Computing (HPC) clusters or the cloud.  It
    is thus crucial to quantify the performance of these applications
    on these platforms. The goals include determining which type of hardware/software
    stacks are best suited to different application classes, as well as
    understanding the limitations of current algorithms, designs and
    technologies. Unfortunately, performance
        studies that rely on real-world experiments on compute platforms
        face several difficulties (high operational costs, labor-intensive experimental setups,
        shared platforms with dynamic loads that hinder reproducibility of results) and have several shortcomings
        (experiments are limited to the available platform/software configurations, which precludes
        the exploration of hypothetical scenarios). 
    Simulation addresses these concerns by providing models
    and abstractions for the performance of computer hardware, such as
    CPU, network and storage. As a result, simulations provide a
    cost-effective, fast, convenient and reproducible way to evaluate
    application performance on arbitrary platform configurations. It thus comes
        as no surprise that a large number of simulation frameworks have been developed and used 
        for research and development in the field of parallel and distributed computing~\cite{ optorsim, gridsim, groudsim, cloudsim, nunez2012simcan,nunez2012icancloud, mdcsim, dissect_cf, cloudnetsimplusplus, fognetsimplusplus, casanova2014simgrid, ROSS, casanova2020fgcs}.

        Page caching is an ubiquitous technique for mitigating the I/O bottleneck.
        As such, it is necessary to model it
        when simulating data-intensive applications.
        While existing simulation frameworks of parallel and distributed computing
        systems  capture many relevant features of hardware/software stacks, 
        most lack the ability to simulate page cache with enough details to capture key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Some simulators, such as the one in~\cite{xu2018saving}, do capture such features
        but are domain-specific.

@@

    We separate our simulation model in two components, the I/O
    Controller and the Memory Manager, which together simulate
    file reads and writes (Figure~\ref{fig:interaction}).
    To read or write a file chunk, a simulated application sends a
    request to the I/O Controller. The I/O Controller interacts as needed with
    the Memory Manager to free memory through flushing or eviction,
    and to read or write cached data. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction, and reads or writes to disk when necessary.
    In case the writethrough strategy is used, the I/O Controller directly writes to disk, 
    cache is flushed if needed and written data is added to page cache.

@@

    The Memory Manager simulates two parallel threads: the main one
    implements flushing, eviction, and cached I/Os synchronously, whereas
    the second one, which operates in the background, periodically searches for
    expired dirty data in LRU lists and flushes this data to disk. We
    rely on the existing SimGrid storage simulation models~\cite{lebre2015} to simulate disk and
    memory, characterized by their storage capacity, read and write
    bandwidths, and latency. These models account for
    bandwidth sharing between concurrent memory or disk accesses.

@@

    The main simulated thread in the Memory Manager can flush or evict data from the
    memory cache. The data flushing simulation
    function takes the amount of data to flush as parameter. While
    this amount is not reached and dirty
    blocks remain in cache, this function traverses the sorted
    inactive list, then the sorted active list, and writes the
    least recently used dirty block to disk, having set its dirty
    flag to 0. In case the amount of data to flush requires that a
    block be partially flushed, the block is split into two blocks,
    one that is flushed and one that remains dirty. The time needed
    to flush data to disk is simulated by the storage model.

    The cache eviction simulation also runs in
    the main thread. It frees up the page cache by traversing and deleting
    least recently used clean data blocks in the inactive list.
    In case the amount evictable memory in the inactive list is insufficient 
    (the inactive list is empty or a file being opened is residing in the inactive list), 
    blocks in the active list are considered for eviction. 
    The amount of data to evict is passed as a parameter and data blocks are deleted
    from the lists until the evicted data reaches the required amount,
    or until there is no clean block left in the list.
    If the last evicted block does not have to be entirely evicted, the block is split into two blocks,
    and only one of them is evicted.
    The overhead of the cache eviction algorithm is not part of the simulated time
    since cache eviction time is negligible in real systems. % \tristan{a reference would be welcome}.

@@

        \end{algorithmic}
    \end{algorithm}
    Algorithm~\ref{alg:write} describes how chunk writes in
    the I/O Controller are simulated.
    % Data is written to cache until the amount of dirty data in the cache
    % exceeds the dirty ratio times the amount of available memory. Data is then written to disk.
    The algorithm initially checks the  amount of dirty data that
    can be written given the dirty ratio (line 5).
    If this amount is greater than 0, the Memory Manager is requested to evict
    data from cache if necessary (line 7).
    After eviction, the amount of data that can be written to
    page cache is calculated (line 8), and a cache write is simulated (line 9).
    If the dirty threshold is reached and there is still data to write,
    the remaining data is written to cache in a loop
    where we repeatedly flush and evict data from the cache (line 12-18).

@@

        Our experiments compare real executions with our Python prototype,
        with the original \wrench simulator, and with our \wrench-cache
        extension. Executions included single-threaded and multi-threaded
        applications, accessing data on local and network file systems. We
        used two applications: a synthetic one, created to evaluate the
        simulation model, and a real one, representative of neuroimaging
        data processing.

@@

        \begin{table}[b]
            \centering
            \begin{tabularx}{0.8\columnwidth}{c>{\centering\arraybackslash}X}
            \toprule
                Input size (GB)  & CPU time (s)\\
            \midrule
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \bottomrule
            \end{tabularx}
            \caption{Synthetic application parameters}
            \label{table:cputime}
            \end{table}
        The synthetic application, implemented in C, consists of three single-core,
        sequential tasks where each task reads the file produced by the
        previous task, increments every byte of this file to emulate real
        processing, and writes the resulting data to disk. Files are
        numbered by ascending access times (File 1 is the file read by Task 1, etc).
         The anonymous memory used by the application
        is released after each task, which we also simulate in the Python
        prototype and in \wrench-cache. As our focus is on I/O rather than compute, we measured
        application task CPU times on a cluster node
        (Table~\ref{table:cputime}), and used these durations in our
        simulations. For the Python prototype, we injected CPU times
        directly into the simulation. For \wrench and \wrench-cache, we
        determined the corresponding number of flops on a 1~Gflops CPU
        and used these values in the simulation. The simulated
        platform and application are available at
        commit \href{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}{ec6b43561b}.

        We used the synthetic application in three experiments. In the
        first one (\textit{Exp~1}), we run \emph{a single instance} of
        the application on a single cluster node, with different input file
        sizes (20~GB, 50~GB, 75~GB, 100~GB), and with all I/Os directed to
        the same local disk.
        In the second experiment (\textit{Exp~2}), we run
        \emph{concurrent} instances of the application on a single node,
        all application instances operating on different files of size 3~GB
        stored in the same local disk. We varied the number of concurrent
        application instances from 1 to 32 since cluster nodes have 32 CPU
        cores.
        In the third experiment (\textit{Exp~3}), we use the same
        configuration as the previous one, albeit reading and writing
        on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
        another compute node. NFS client and server read caches were enabled, however, as is commonly configured in HPC
        environments to avoid data loss, there was no client write cache
        and the server cache was configured as writethrough instead of
        writeback.
        Therefore, all the writes happened at disk bandwidth, but
        reads could benefit from cache hits.

@@

        We used the real application in the fourth experiment
        (\textit{Exp~4}), which runs on a single cluster node 
        using a single local disk, to process data from
        participant 0027430 in the dataset of the Max Planck Institute for
        Human Cognitive and Brain Sciences available at
        \url{http://dx.doi.org/10.15387/fcp_indi.corr.mpg1}, leading to the
        parameters in Table~\ref{table:nighres_stats}.

@@

        The page cache simulation model drastically reduces I/O simulation
        errors in each application task (Figure~\ref{fig:single_error}). Errors were reduced from an average
        of 328\% in the original \wrench to 36\% in the Python prototype and
        32\% in \wrench-cache. Unsurprisingly, the original \wrench simulator
        significantly overestimated read and write times, due to the lack
        of page cache simulation. Consistently, the first read was not impacted
        as it only involved uncached data. Results with files of 50~GB and 75~GB
        showed similar behaviors and are not reported for brevity.

        \wrench simulation errors are substantially lower with 100~GB
        files than with 20~GB files, due to the fact that part of the
        100~GB file needs to be read and written to disk as it does not fit in cache. Disk models are
        identical in WRENCH and WRENCH-cache.

        Simulated memory profiles are highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). With 20~GB files, memory profiles almost exactly match the
        real ones, although for some reason dirty data seems to be flushing faster in real
        life than in simulation, a behavior also
        observed with 100~GB files. With 100~GB files, used memory reaches
        total memory during the first write, triggering dirty data
        flushing, and drops back to cached memory when application tasks
        release anonymous memory. Simulated cached memory is highly
        consistent with real values. In all cases, dirty data remains under the dirty ratio as
        expected. The Python prototype and \wrench-cache exhibit nearly
        identical memory profiles, which reinforces the confidence in our
        implementations.

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}), showing that the
        previous observations were not incidental.

@@

            The page cache model notably reduces \wrench's simulation error
            for concurrent applications executed with local I/Os
            (Figure~\ref{fig:multi_local}). For reads (Figure~\ref{fig:multi_local}-left), \wrench-cache
            slightly overestimates runtime, due to the aforementioned discrepancy between
            simulated and real read bandwidths.
            For writes (Figure~\ref{fig:multi_local}-right), \wrench-cache results show a
            plateau similar to the one observed in the real
            execution between 1 and 10 concurrent executions. The plateau marks the limit beyond which the page cache is
            saturated with dirty data and needed flushing.

@@

            Page cache simulation  reduces simulation error significantly for
            NFS storage as well (Figure~\ref{fig:multi_nfs}). This
            is manifested only for reads, as the NFS server uses writethrough rather than writeback caching.
            Both \wrench and \wrench-cache
            underestimate write times due to the aforementioned discrepancy between
            simulated and real bandwidths. For reads,
            this discrepancy only impacts the results beyond 22
            applications since before this threshold most reads result in cache
            hits.

@@

        Similar to the synthetic application, simulation error for the real application is
        substantially reduced by the \wrench-cache simulator compared to
        \wrench (Figure~\ref{fig:nighres}). On average, errors are reduced
        from 337~\% in \wrench to 47~\% in \wrench-cache. 
        The first read happens entirely from disk and is thus 
        very accurately simulated by both \wrench and \wrench-cache.

@@

        As is the case for \wrench, simulation time with \wrench-cache scales
        linearly with the number of concurrent applications
        (Figure~\ref{fig:multi_time}, p \textless $10^{-17}$). However, the page
        cache model substantially increases simulation time by
        application, as can be seen by comparing regression slopes in
        Figure~\ref{fig:multi_time}. Interestingly, \wrench-cache is faster with 
        NFS I/Os than with local I/Os, most likely due to the use of writethrough
        cache in NFS, which bypasses flushing operations.

        \begin{figure}
            \centering
            \includegraphics[width=0.95\linewidth]{result/multi/figures/simulation_time.pdf}
            \caption{Simulation time comparison. \wrench-cache scales
            linearly with the number of concurrent applications, albeit
            with a higher overhead than \wrench.}
            \label{fig:multi_time}
            \end{figure}

    \section{Conclusion}
    \label{discussion}
    We designed a model of the Linux page cache and implemented it in the
    \simgrid-based \wrench simulation framework to simulate the execution
    of distributed applications.
    Evaluation results show that our model improves simulation accuracy
    substantially, reducing absolute relative simulation errors by up to
    9$\times$ (see results of the single-threaded experiment). The
    availability of asymmetrical disk bandwidths in the forthcoming
    \simgrid release will further improve these results.
    Our page cache model is publicly available in the \wrench GitHub
    repository.

    Page cache simulation can be instrumental in a number of studies. For
    instance, it is now common for HPC clusters to run applications in
    Linux control groups (cgroups), where resource consumption is limited,
    including memory and therefore page cache usage. Using our simulator,
    it would be possible to study the interaction between memory allocation
    and I/O performance, for instance to improve scheduling algorithms or
    avoid page cache starvation~\cite{zhuang2017}. Our simulator could also
    be leveraged to evaluate solutions that reduce the impact of network
    file transfers on distributed applications, such as burst
    buffers~\cite{ferreiradasilva-fgcs-bb-2019}, hierarchical file
    systems~\cite{islam2015triple}, active storage~\cite{5496981}, or
    specific hardware architectures~\cite{hayot2020performance}.

    Not all I/O behaviors are captured by currently available simulation models,
    including the one developed in this work, 
    which limits simulation accuracy.
    Relevant extensions to this work include more
    accurate descriptions of anonymous memory usage in applications, 
    which strongly affects I/O times through writeback caching. File access patterns
    might also be worth including in the simulation models,
    as they directly affect page cache content.

@@

\clearpage
\bibliographystyle{IEEEtran}
\bibliography{citation}

>>>>

    The emergence of Big Data in recent years has resulted in a growing
    need for efficient data processing solutions. While infrastructures
    with sufficient compute power are available,
    the I/O bottleneck remains. The Linux page cache is an efficient
    approach to reduce I/O overheads, but few
    experimental studies of its interactions with Big Data applications exist,
    partly due to limitations of
    real-world experiments. Simulation is a popular approach to address
    these issues, however, existing simulation frameworks do not simulate
    page caching fully, or even at all.  As a result, simulation-based
    performance studies of data-intensive applications can lead to misleading
    results and inaccurate conclusions.

    In this paper, we propose an I/O simulation model that captures
    the key features of the Linux page cache. We have implemented this model
    as part of the \wrench workflow simulation framework, which itself
    builds on the popular \simgrid distributed systems simulation
    framework. Our model and its implementation enable the simulation
    of both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems. We evaluate the accuracy of our model in different
    conditions, including sequential and concurrent applications, as
    well as local and remote I/Os. We find that our page cache model
    reduces the simulation error by up to an order of magnitude when
    compared to state-of-the-art, cacheless simulations. Our model is publicly available in the 
    \wrench framework, making it usable in a wide range of simulation studies.

@@

    The Linux page cache plays an important role in
    reducing filesystem data transfer times. With the page cache, previously
    read data can be re-read directly from memory, and written data can be written to
    memory before being asynchronously flushed to disk, resulting in improved I/O performance
    on slower storage devices. The performance improvements
    depend on many factors including the total amount of memory,
    the amount of data being written (i.e., dirty data), and the amount of memory available for
    written data. All these factors are important when determining the impact of I/O on
    application performance, particularly for data-intensive applications.
 % Swapped the order the two paragraphs and merged them into one
    The number of data-intensive applications has been steadily rising as a result of
    open-data and data sharing initiatives. Due to the sheer size of the data being
    processed, these applications must be executed on large-scale infrastructures
    such as High Performance Computing (HPC) clusters or the cloud.  It
    is thus crucial to quantify the performance of these applications
    on these platforms. The goals include determining which type of hardware/software
    stacks are best suited to different application classes, as well as
    understanding the limitations of current algorithms, designs and
    technologies. Unfortunately, performance
    studies that rely on real-world experiments on compute platforms
    face several difficulties (high operational costs, labor-intensive experimental setups,
    shared platforms with dynamic loads that hinder reproducibility of results) and have several shortcomings
    (experiments are limited to the available platform/software configurations, which precludes
    the exploration of hypothetical scenarios). 
    Simulation addresses these concerns by providing models
    and abstractions for the performance of computer hardware, such as
    CPU, network and storage. As a result, simulations provide a
    cost-effective, fast, convenient and reproducible way to evaluate
    application performance on arbitrary platform configurations. It thus comes
    as no surprise that a large number of simulation frameworks have been developed and used 
    for research and development in the field of parallel and distributed computing~\cite{ optorsim, gridsim, groudsim, cloudsim, nunez2012simcan,nunez2012icancloud, mdcsim, dissect_cf, cloudnetsimplusplus, fognetsimplusplus, casanova2014simgrid, ROSS, casanova2020fgcs}.

    Page caching is an ubiquitous technique for mitigating the I/O
    bottleneck, reducing disk read and write times by up to an order of
    magnitude. As such, it is necessary to model it
    when simulating data-intensive applications.
    While existing simulation frameworks of parallel and distributed computing
    systems  capture many relevant features of hardware/software stacks, 
    most lack the ability to simulate page cache with enough details to capture key features such
    as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
    Some simulators, such as the one in~\cite{xu2018saving}, do capture such features
    but are domain-specific. In addition, analytical models of the page
    cache have been proposed ~\cite{DBLP:journals/concurrency/InacioD18,
    DBLP:journals/pacmpl/BaoKPS18, DBLP:conf/pldi/GysiGBH19} but their
    integration in simulation frameworks is challenging. A scalable, general and accurate
    page cache simulation model is required to enable reliable simulations of
    data-intensive applications.

@@

    We separate our simulation model in two components, the I/O
    Controller and the Memory Manager, which together simulate
    file reads and writes (Figure~\ref{fig:interaction}).
    To read or write a file chunk, a simulated application sends a
    request to the I/O Controller. The I/O Controller interacts as needed with
    the Memory Manager to free memory through flushing or eviction,
    and to read or write cached data. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction, and reads or writes to disk when necessary.

@@

    The Memory Manager simulates two parallel threads: the main one
    implements flushing, eviction, and cached I/Os synchronously, whereas
    the second one, which operates in the background, periodically searches for
    expired dirty data in LRU lists and flushes this data to disk. We
    rely on the existing SimGrid storage simulation models to simulate disk and
    memory, characterized by their storage capacity, read and write
    bandwidths, and latency. These models account for bandwidth sharing
    between concurrent memory or disk accesses. We refer the reader to~\cite{lebre2015} for a
    detailed description and validation of these models.

@@

    The main simulated thread in the Memory Manager can flush or evict data from the
    memory cache. The data flushing simulation
    function takes the amount of data to flush as parameter. While
    this amount is not reached and dirty
    blocks remain in cache, this function traverses the sorted
    inactive list, then the sorted active list, and writes the
    least recently used dirty block to disk, having set its dirty
    flag to 0. In case the amount of data to flush requires that a
    block be partially flushed, the block is split into two blocks,
    one that is flushed and one that remains dirty. The time needed
    to flush data to disk is simulated by the SimGrid storage model.

    The cache eviction simulation also runs in
    the main thread. It frees up the page cache by traversing and deleting
    least recently used clean data blocks in the inactive list.
    In case the amount of evictable memory in the inactive list is insufficient 
    (the inactive list is empty or only contains blocks that belong to open files), 
    blocks in the active list are considered for eviction. 
    The amount of data to evict is passed as a parameter and data blocks are deleted
    from the lists until the evicted data reaches the required amount,
    or until there is no clean block left in the lists.
    If the last evicted block does not have to be entirely evicted, the block is split into two blocks,
    and only one of them is evicted.
    The overhead of the cache eviction algorithm is not part of the simulated time
    since cache eviction time is negligible in real systems. % \tristan{a reference would be welcome}.

@@

        \end{algorithmic}
    \end{algorithm}
    Algorithm~\ref{alg:write} describes how chunk writes in
    the I/O Controller are simulated.
    % Data is written to cache until the amount of dirty data in the cache
    % exceeds the dirty ratio times the amount of available memory. Data is then written to disk.
    The algorithm initially checks the  amount of dirty data that
    can be written given the dirty ratio (line 5).
    If this amount is greater than 0, the Memory Manager is requested to evict
    data from cache if necessary to accomodate the chunk size (line 7).
    After eviction, the amount of data that can be written to
    page cache is calculated (line 8), and a cache write is simulated (line 9).
    If the dirty threshold is reached and there is still data to write,
    the remaining data is written to cache in a loop
    where we repeatedly flush and evict data from the cache (line 12-18).

@@

        Our experiments compared real executions with our Python prototype,
        with the original \wrench simulator, and with our \wrench-cache
        extension. Executions included single-threaded and multi-threaded
        applications, accessing data on local and network file systems. We
        used two applications: a synthetic one, created to evaluate the
        simulation model, and a real one, representative of neuroimaging
        data processing.

@@

        \begin{table}[b]
            \centering
            \begin{tabularx}{0.8\columnwidth}{c>{\centering\arraybackslash}X}
            \toprule
                Input size (GB)  & CPU time (s)\\
            \midrule
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \bottomrule
            \end{tabularx}
            \caption{Synthetic application parameters}
            \label{table:cputime}
            \end{table}
        The synthetic application, implemented in C, consists of three single-core,
        sequential tasks where each task reads the file produced by the
        previous task, increments every byte of this file to emulate real
        processing, and writes the resulting data to disk. Files are
        numbered by ascending access times (File 1 is the file read by Task 1, etc).
         The anonymous memory used by the application
        is released after each task, which we also simulate in the Python
        prototype and in \wrench-cache. As our focus is on I/O rather than compute, we measured
        application task CPU times on a cluster node
        (Table~\ref{table:cputime}), and used these durations in our
        simulations. For the Python prototype, we injected CPU times
        directly into the simulation. For \wrench and \wrench-cache, we
        determined the corresponding number of flops on a 1~Gflops CPU
        and used these values in the simulation. The simulated
        platform and application are available in WRENCH at
        commit \href{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}{ec6b43561b}.

        We used the synthetic application in three experiments. In the
        first one (\textit{Exp~1}), we ran \emph{a single instance} of
        the application on a single cluster node, with different input file
        sizes (20~GB, 50~GB, 75~GB, 100~GB), and with all I/Os directed to
        the same local disk.
        In the second experiment (\textit{Exp~2}), we ran
        \emph{concurrent} instances of the application on a single node,
        all application instances operating on different files of size 3~GB
        stored in the same local disk. We varied the number of concurrent
        application instances from 1 to 32 since cluster nodes have 32 CPU
        cores.
        In the third experiment (\textit{Exp~3}), we used the same
        configuration as the previous one, albeit reading and writing
        on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
        another compute node. NFS client and server read caches were enabled, however, as is commonly configured in HPC
        environments to avoid data loss in netework file systems, there was no client write cache
        and the server cache was configured as writethrough instead of
        writeback.
        Therefore, all the writes happened at disk bandwidth, but
        reads could benefit from cache hits.

@@

        In the fourth experiment
        (\textit{Exp~4}), we ran the real application on a single cluster node, 
        using a single local disk, to process data from
        participant 0027430 in the dataset of the Max Planck Institute for
        Human Cognitive and Brain Sciences available at
        \url{http://dx.doi.org/10.15387/fcp_indi.corr.mpg1}. The execution parameters are reported
        in Table~\ref{table:nighres_stats}.

@@

        The page cache simulation model drastically reduced I/O simulation
        errors in each application task (Figure~\ref{fig:single_error}). Errors were reduced from an average
        of 328\% in the original \wrench to 36\% in the Python prototype and
        32\% in \wrench-cache. Unsurprisingly, the original \wrench simulator
        significantly overestimated read and write times, due to the lack
        of page cache simulation. Consistently, the first read was not impacted
        as it only involved uncached data. Results with files of 50~GB and 75~GB
        showed similar behaviors and are not reported for brevity.

        \wrench simulation errors were substantially lower with 100~GB
        files than with 20~GB files, due to the fact that part of the
        100~GB file needed to be read and written to disk as it did not fit in cache. Disk models are
        identical in WRENCH and WRENCH-cache.

        Simulated memory profiles were highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). With 20~GB files, memory profiles almost exactly matched the
        real ones, although for some reason dirty data seemed to be flushing faster in real
        life than in simulation, a behavior also
        observed with 100~GB files. With 100~GB files, used memory reached
        total memory during the first write, triggering dirty data
        flushing, and droped back to cached memory when application tasks
        released anonymous memory. Simulated cached memory was highly
        consistent with real values. In all cases, dirty data remained under the dirty ratio as
        expected. The Python prototype and \wrench-cache exhibited nearly
        identical memory profiles, which reinforces the confidence in our
        implementations.

        The content of the simulated memory cache was also highly consistent
        with reality (Figure~\ref{fig:single_cache}), showing that the
        previous observations were not incidental.

@@

            The page cache model notably reduced \wrench's simulation error
            for concurrent applications executed with local I/Os
            (Figure~\ref{fig:multi_local}). For reads (Figure~\ref{fig:multi_local}-left), \wrench-cache still
            slightly overestimates runtime, due to the aforementioned discrepancy between
            simulated and real read bandwidths.
            For writes (Figure~\ref{fig:multi_local}-right), \wrench-cache results show a
            plateau similar to the one observed in the real
            execution between 1 and 10 concurrent executions. The plateau marks the limit beyond which the page cache was
            saturated with dirty data and needed flushing.

@@

            Page cache simulation  reduced simulation error significantly for
            NFS storage as well (Figure~\ref{fig:multi_nfs}). This
            was manifested only for reads, as the NFS server used writethrough rather than writeback caching.
            Both \wrench and \wrench-cache
            underestimated write times due to the aforementioned discrepancy between
            simulated and real bandwidths. For reads,
            this discrepancy only impacted the results beyond 22
            applications since before this threshold most reads resulted in cache
            hits.

@@

        Similar to the synthetic application, simulation error for the real application was
        substantially reduced by the \wrench-cache simulator compared to
        \wrench (Figure~\ref{fig:nighres}). On average, errors were reduced
        from 337~\% in \wrench to 47~\% in \wrench-cache. 
        The first read happened entirely from disk and was thus 
        very accurately simulated by both \wrench and \wrench-cache.

@@

        Simulation time with \wrench-cache scaled
        linearly with the number of concurrent applications, as was the case for \wrench
        (Figure~\ref{fig:multi_time}, p \textless $10^{-16}$). However, the page
        cache model substantially increased simulation time by
        application, as can be seen by comparing regression slopes in
        Figure~\ref{fig:multi_time}. Interestingly, \wrench-cache was faster with 
        NFS I/Os than with local I/Os, most likely due to the use of writethrough
        cache in NFS, which bypasses flushing operations.

        \begin{figure}
            \centering
            \includegraphics[width=0.95\linewidth]{result/multi/figures/simulation_time.pdf}
            \caption{Simulation time comparison. \wrench-cache scaled
            linearly with the number of concurrent applications, albeit
            with a higher overhead than \wrench.}
            \label{fig:multi_time}
            \end{figure}

    \section{Conclusion}
    \label{discussion}
    We designed a model of the Linux page cache and implemented it in the
    \simgrid-based \wrench simulation framework to simulate the execution
    of distributed applications.
    Evaluation results show that our model improves simulation accuracy
    substantially, reducing relative simulation errors by up to
    10$\times$. The
    availability of asymmetrical disk bandwidths in the forthcoming
    \simgrid release will further improve these results.
    Our page cache model is publicly available in the \wrench GitHub
    repository and can therefore be applied transparently to simulate data-intensive applications.

    Page cache simulation can be instrumental in a number of studies. For
    instance, it is now common for HPC clusters to run applications in
    Linux control groups (cgroups), where resource consumption is limited,
    including memory and therefore page cache usage. Using our simulator,
    it would be possible to study the interaction between memory allocation
    and I/O performance, for instance to improve scheduling algorithms or
    avoid page cache starvation~\cite{zhuang2017}. Our simulator could also
    be leveraged to evaluate solutions that reduce the impact of network
    file transfers on distributed applications, such as burst
    buffers~\cite{ferreiradasilva-fgcs-bb-2019}, hierarchical file
    systems~\cite{islam2015triple}, active storage~\cite{5496981}, or
    specific hardware architectures~\cite{hayot2020performance}. We expect
    page cache simulation to drastically reduce the simulation error in all
    these scenarios, with substantial consequences on study conclusions.

    Relevant extensions to this work include more
    accurate descriptions of anonymous memory usage in applications, 
    which strongly affects I/O times through writeback caching. File access patterns
    might also be worth including in the simulation models,
    as they directly affect page cache content.

@@

\bibliographystyle{IEEEtran}
\bibliography{citation}
